Manual classification of tumor cells

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/01-marker-genes.Rmd
account for no marker gene expression

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) 
```

@@

```{r}
ggplot(umap_df, aes(x = gene_expression, fill = gene_symbol)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

```{r}
umap_df <- umap_df |> 
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
  dplyr::ungroup()
```

```{r}
ggplot(umap_df, aes(x = transformed_gene_expression, fill = gene_symbol)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

```{r}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- umap_df |> 
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()

@@

```{r}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- marker_sum_exp |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))

@@

```{r}
# add cell types from sum classification to plotting df 
celltype_df <- umap_df |> 
  dplyr::left_join(marker_sum_exp)

@@

```{r}
# add in tumor/normal classification to colData of SCE 
celltypes_only_df <- marker_sum_exp |> 
  dplyr::select(barcodes, sum_classification)

@@

```{r}
# pull out top markers for tumor cells 
tumor_markers <- markers$Tumor |> 
  as.data.frame() |> 
  tibble::rownames_to_column("ensembl_gene_id") |> 
  # join to get gene symbols to print out later
  dplyr::left_join(marker_genes_df) |>
  # just take those with FDR < 0.05 and logFC > 1.5
  dplyr::filter(FDR < 0.05, summary.logFC > 1.5) |> 
  dplyr::arrange(FDR)

@@

```{r, fig.height=10, fig.width=10}
marker_plot_df <- celltype_df |> 
  dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |> 
  dplyr::select(gene_symbol, gene_expresed, sum_classification)

@@

```{r}
# get all assigned cell types in one dataframe 
celltypes_df <- as.data.frame(colData(sce)) |> 
  dplyr::select(barcodes, singler_celltype_annotation, cellassign_celltype_annotation) |> 
  dplyr::left_join(marker_sum_exp)

@@

```{r, fig.height=10, fig.width=7}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))

@@

```{r}
# plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign 
# use already identified normal/tumor cell types 
if(!is.null(ref_cells_df)){

@@

```{r}
# get a dataframe with barcodes, sum_classification and cluster_classification 
celltypes_df <- celltypes_df |> 
  dplyr::select(
    barcodes,
    marker_gene_classification = sum_classification
  )

readr::write_tsv(celltypes_df, output_classifications_file)
```

>>>>

```{r}
if(sum(umap_df$gene_expression) == 0){
  has_marker_gene_exp <- FALSE
} else {
  has_marker_gene_exp <- TRUE
}
```

```{r, eval=has_marker_gene_exp}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) 
```

@@

```{r, eval=has_marker_gene_exp}
ggplot(umap_df, aes(x = gene_expression, fill = gene_symbol)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

```{r, eval=has_marker_gene_exp}
umap_df <- umap_df |> 
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
  dplyr::ungroup()
```

```{r, eval=has_marker_gene_exp}
ggplot(umap_df, aes(x = transformed_gene_expression, fill = gene_symbol)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

```{r, eval=has_marker_gene_exp}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- umap_df |> 
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()

@@

```{r, eval=has_marker_gene_exp}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- marker_sum_exp |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))

@@

```{r, eval=has_marker_gene_exp}
# add cell types from sum classification to plotting df 
celltype_df <- umap_df |> 
  dplyr::left_join(marker_sum_exp)

@@

```{r, eval=has_marker_gene_exp}
# add in tumor/normal classification to colData of SCE 
celltypes_only_df <- marker_sum_exp |> 
  dplyr::select(barcodes, sum_classification)

@@

```{r, eval=has_marker_gene_exp}
# pull out top markers for tumor cells 
tumor_markers <- markers$Tumor |> 
  as.data.frame() |> 
  tibble::rownames_to_column("ensembl_gene_id") |> 
  # join to get gene symbols to print out later
  dplyr::left_join(marker_genes_df) |>
  # just take those with FDR < 0.05 and logFC > 1.5
  dplyr::filter(FDR < 0.05, summary.logFC > 1.5) |> 
  dplyr::arrange(FDR)

@@

```{r, fig.height=10, fig.width=10, eval=has_marker_gene_exp}
marker_plot_df <- celltype_df |> 
  dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |> 
  dplyr::select(gene_symbol, gene_expresed, sum_classification)

@@

```{r, eval=has_marker_gene_exp}
# get all assigned cell types in one dataframe 
celltypes_df <- as.data.frame(colData(sce)) |> 
  dplyr::select(barcodes, singler_celltype_annotation, cellassign_celltype_annotation) |> 
  dplyr::left_join(marker_sum_exp)

@@

```{r, fig.height=10, fig.width=7, eval=has_marker_gene_exp}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))

@@

```{r, eval=has_marker_gene_exp}
# plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign 
# use already identified normal/tumor cell types 
if(!is.null(ref_cells_df)){

@@

```{r}
if(has_marker_gene_exp){
  # get a dataframe with barcodes, sum_classification and cluster_classification 
  celltypes_df <- celltypes_df |> 
    dplyr::select(
      barcodes,
      marker_gene_classification = sum_classification
    )

  readr::write_tsv(celltypes_df, output_classifications_file)
} else{
  # save marker gene classification as NA if no expression 
  celltypes_df <- data.frame(
    barcodes = umap_df$barcodes,
    marker_gene_classification = NA
  )

  readr::write_tsv(celltypes_df, output_classifications_file)
}
```
